\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-11-05 lang-design.tex
trying out a name for the language design and impl

We use a plugin to the Coq proof assistant to demonstrate the key
ingredients of our design,
though we believe our design could be adapted to other proof assistants such as Lean.

@@

\cref{fig:stlc-mechanized} shows how STLC and its extension with
fixpoints can be mechanized using our plugin, in a style envisioned in
\cref{fig:stlc-nonmechanized}.
The base family "STLC" hosts the STLC metatheories, from abstract syntax
to the type-safety theorem.
Family "STLCFix", derived from "STLC", makes adjustments as needed by a fixpoints extension:
it adds new constructors to the inductive types (\lsti{FInductive}) and
adds new cases to the recursive functions (\lsti{FRecursion}) and
induction proofs (\lsti{FInduction}).
Existing constructors and cases, as well as those definitions and
theorems that need no adjustments ("ty", "env", "empty", "steps", and
"typesafe"), are automatically inherited and reused.
In particular, executing the last command, \lsti{Check STLCFix.typesafe},
displays the type-safety theorem of the fixpoints extension.

@@

A key observation is that although late binding
prevents \emph{definitional equality} on \lsti{subst}, it does not
affect \emph{propositional equality}.
That is, \lsti{\forall x t, subst tm_unit x t = tm_unit},
as a proposition (\lsti{Prop}), should still hold.
After all, how \lsti{subst} is defined on \lsti{tm_unit} does not vary
from a base family to a derived family.\EDJ{I think the reviewer might ask : what if the subst late-bound tm_unit case/why the subst cases cannot be late-bound? it not necessary return tm_unit in those case. 
\\ What the plugin generates is actually subst tm_unit x t = subst.tm_unit. In other words, it points back to the handler instead of the concrete constructor. That is one of the advantage of looking FRecursion as a family.}
%
Based on this insight, our plugin automatically generates
propositional equalities for each recursive function defined with \lsti{FRecursion},
making them available as axioms for use by the rest of the current family.
The plugin also provides a tactic \lsti{fsimpl} that enables, for instance,
simplifying \lsti{subst tm_unit x t'} to \lsti{tm_unit}.
Under the hood, \lsti{fsimpl} works by rewriting using the postulated
propositional equalities.

@@

Coq distinguishes \emph{opaque} definitions from \emph{transparent} ones.
Our plugin supports the overriding of opaque definitions, which
include most proofs.
It is safe to override opaque definitions, because Coq will never attempt to unfold them.
For transparent definitions, the common case is that the programmer does
not want to override them.
In \cref{fig:stlc-mechanized}, "env", "empty", "subst", and "steps" are
transparent---that is, they are not defined with \lsti{Qed}.
Our plugin treats transparent definitions as non-overridable by default.
Thus, the definitional equalities on "env", "empty", and "steps",
as well as the propositional equality on "subst", are available to the
type checker for type-checking the families.

Our plugin does allow the overriding of transparent definitions that are
explicitly marked as \lsti{Overridable} by the programmer.
Overriding is made safe by requiring that when an overridable field is
overridden, code whose type checking involves unfolding that field
should be overridden too.
We expect this feature to be used only occasionally.
Our case studies contain uses of overriding (\cref{sec:casestudies}).

@@

Mixin composition is a form of multiple inheritance, which may lead to
name conflicts in general.
Our plugin requires the programmer to resolve conflicts by
explicitly overriding conflicted fields.

@@

To provide a streamlined programming experience (\ref{chg:tooling}),
our plugin offers two tactics, \lsti{finjection} and \lsti{fdiscriminate}.
For example, in a proof state that contains a manifestly false assumption
\lsti{H : tm_var "x" = tm_abs "y" t}, the programmer can use
\lsti{fdiscriminate H} to obtain \lsti{False} and thus discharge
the current goal, just as they would with \lsti{discriminate H} if "tm"
were not extensible.

>>>>

We present the key ingredients of our design as an extension to the Coq
proof assistant,
though we believe it could be adapted to other proof assistants such as Lean.
We call our design and implementation \Lang (\underline{f}amily
\underline{\smash{po}}lymorphism for a \underline{\smash{p}}roof assistant).
In this section, we focus on the language design of \Lang.
\cref{sec:coqimpl} describes its implementation as a Coq plugin.

@@

\cref{fig:stlc-mechanized} shows how STLC and its extension with
fixpoints can be mechanized using \Lang, in a style envisioned in
\cref{fig:stlc-nonmechanized}.
The base family "STLC" hosts the STLC metatheories, from abstract syntax
to the type-safety theorem.
Family "STLCFix", derived from "STLC", makes adjustments as needed by a fixpoints extension:
it adds new constructors to the inductive types (\lsti{FInductive}) and
adds new cases to the recursive functions (\lsti{FRecursion}) and
induction proofs (\lsti{FInduction}).
Existing constructors and cases, as well as those definitions and
theorems that need no adjustments ("ty", "env", "empty", "steps", and
"typesafe"), are automatically inherited and reused.
In particular, executing the last command, \lsti{Check STLCFix.typesafe},
displays the type-safety theorem of the fixpoints extension.

@@

A key observation is that although late binding
prevents \emph{definitional equality} on \lsti{subst}, it does not
affect \emph{propositional equality}.
That is, \lsti{\forall x t, subst tm_unit x t = tm_unit},
as a proposition (\lsti{Prop}), should still hold.
After all, how \lsti{subst} is defined on \lsti{tm_unit} does not vary
from a base family to a derived family.\EDJ{I think the reviewer might ask : what if the subst late-bound tm_unit case/why the subst cases cannot be late-bound? it not necessary return tm_unit in those case. 
\\ What the plugin generates is actually subst tm_unit x t = subst.tm_unit. In other words, it points back to the handler instead of the concrete constructor. That is one of the advantage of looking FRecursion as a family.}
%
Based on this insight, \Lang automatically generates
propositional equalities for each recursive function defined with \lsti{FRecursion},
making them available as axioms for use by the rest of the current family.
\Lang also provides a tactic \lsti{fsimpl} that enables, for instance,
simplifying \lsti{subst tm_unit x t'} to \lsti{tm_unit}.
Under the hood, \lsti{fsimpl} works by rewriting using the postulated
propositional equalities.

@@

Coq distinguishes \emph{opaque} definitions from \emph{transparent} ones.
\Lang supports the overriding of opaque definitions, which
include most proofs.
It is safe to override opaque definitions, because Coq will never attempt to unfold them.
For transparent definitions, the common case is that the programmer does
not want to override them.
In \cref{fig:stlc-mechanized}, "env", "empty", "subst", and "steps" are
transparent---that is, they are not defined with \lsti{Qed}.
\Lang treats transparent definitions as non-overridable by default.
Thus, the definitional equalities on "env", "empty", and "steps",
as well as the propositional equality on "subst", are available to the
type checker for type-checking the families.

\Lang does allow the overriding of transparent definitions that are
explicitly marked as \lsti{Overridable} by the programmer.
Overriding is made safe by requiring that when an overridable field is
overridden, code whose type checking involves unfolding that field
should be overridden too.
We expect this feature to be used only occasionally.
Our case studies contain uses of overriding (\cref{sec:casestudies}).

@@

Mixin composition is a form of multiple inheritance, which may lead to
name conflicts in general.
\Lang requires the programmer to resolve conflicts by
explicitly overriding conflicted fields.

@@

To provide a streamlined programming experience (\ref{chg:tooling}),
\Lang offers two tactics, \lsti{finjection} and \lsti{fdiscriminate}.
For example, in a proof state that contains a manifestly false assumption
\lsti{H : tm_var "x" = tm_abs "y" t}, the programmer can use
\lsti{fdiscriminate H} to obtain \lsti{False} and thus discharge
the current goal, just as they would with \lsti{discriminate H} if "tm"
were not extensible.
